\clearpage
\chapter*{Colophon: archiving this note}
\addcontentsline{toc}{chapter}{Colophon: archiving this note}
\label{ch:colophon}

This note is a worked example of its own subject. Its \LaTeX{} source is archived in
Software Heritage, and the intrinsic identifier of the archived source directory is:

\begin{center}
\ifdefempty{\thisswhid}{%
  \emph{(stamped at release time --- see ``How this identifier is produced'' below)}%
}{%
  \swhid{\thisswhid}%
}
\end{center}

\paragraph{Why this identifier is not written into the source.}
A \textsc{swhid} is a cryptographic hash of the content it names. Writing a document's own
\textsc{swhid} \emph{into} that document would change the document's bytes, and therefore
change its hash --- and you cannot, in any case, embed a digest you have not computed yet, and
cannot compute until the bytes are fixed. This is not a limitation of Software Heritage; it is what
makes the identifier
trustworthy: the name is forced to follow the bytes, so it cannot be faked. The source is
therefore archived first, and the resulting identifier is stamped only into \emph{this
rendered PDF}, which is not part of the hashed source tree.

\paragraph{How this identifier is produced.}
After the source is committed and archived (via Save Code Now), the qualified directory
\textsc{swhid} is read from the archive's \emph{Permalinks} tab and stamped into a
non-committed PDF:
\begin{center}
\texttt{make swhid SWHID='swh:1:dir:...;origin=...;visit=...'}
\end{center}
A plain \texttt{make} deliberately leaves the slot empty: the committed source never contains
its own identifier, so the source's \textsc{swhid} stays well-defined and stable.

\begin{handson}[Hands-on (capstone): verify this very note]
You can recompute the identifier above and confirm it yourself.
\begin{enumerate}[nosep,leftmargin=*]
  \item Export the \emph{committed} source (tracked files only, no build artefacts):\\
        \texttt{git archive HEAD | (mkdir -p /tmp/note \&\& tar -x -C /tmp/note)}
  \item Compute the directory \textsc{swhid} with the Rust reference tool (\texttt{cargo install
        swhid}; see Appendix~\ref{app:swhid-tools}): \texttt{swhid dir /tmp/note}.
  \item Compare its core \swhid{swh:1:dir:<hash>} with the one printed above (ignore the
        \texttt{origin}/\texttt{visit} qualifiers). A match means you have verified the
        integrity of the document you are reading.
  \item \emph{Bonus (git-compatibility):} the note's \emph{revision} identifier is just
        \swhid{swh:1:rev:} prepended to \texttt{git rev-parse HEAD} --- a Software Heritage
        revision \textsc{swhid} \emph{is} the git commit hash.
\end{enumerate}
\end{handson}
