% The Source Code of Science — a doctoral course note.
% Build with:  latexmk -xelatex main.tex   (or: make)
\documentclass[11pt,oneside]{report}

\input{preamble}

\begin{document}

\swhtitlepage

% --- Abstract / how to read this note ------------------------------------
\chapter*{About this note}
\addcontentsline{toc}{chapter}{About this note}
This note is a doctoral-level course on managing the \emph{source code} of research:
how to archive it for the long term, reference it at the right granularity, describe and
cite it, and use it to make computational results reproducible. It is built on Software
Heritage\footnote{In the interest of transparency: the author helped found Software Heritage,
co-developed the \texttt{biblatex-software} style used here, and contributed to the \textsc{swhid}
standard; the recurring worked example (the Parmap library) is also his. The argument is meant to
stand on its evidence, and the sources cited throughout are independent.} and the \textsc{swhid} ---
an identifier computed from the code itself, defined in Chapter~\ref{ch:identifiers} and now the
international standard ISO/IEC~18670:2025. It addresses
the FAIR principles and then goes \emph{beyond} them, following the CODE-beyond-FAIR roadmap for
research software~\autocite{NatureSD2026}. The acronyms in this paragraph are all explained, in
order, in the chapters that follow; nothing here needs to be known in advance.

The note is meant to be read \emph{and} done: each practical chapter ends with a
\textbf{Hands-on} box. Margin notes flag the key takeaways.

\begin{callout}{In one page (for the busy reader)}
\small
\textbf{The claim.} Source code is a first-class research output, but the habit of depositing a
\texttt{.zip} with a DOI fails it: a DOI certifies a landing page, not the bytes; the deposit can
change, has no granularity, and discards history. The fix is to archive the living repository in
\textbf{Software Heritage} and reference it with a \textbf{\textsc{swhid}} --- an identifier computed
from the content itself, recomputable and verifiable offline, and the international standard
ISO/IEC~18670.

\textbf{The evidence.} The archive holds the development history of more than 28~billion source
files from over 430~million projects. The \textsc{swhid} is already accepted by ACM for the
\emph{Artifacts Available} badge; \emph{eLife} has auto-archived every article's code since 2020;
HAL, IPOL, swMATH, and Episciences are wired in; UNESCO, EOSC, France's National Plan, and the ANR
all point research software at Software Heritage.

\textbf{What to do.}
\emph{Researchers}: archive your code, cite it with a \textsc{swhid} (Chapter~\ref{ch:ardc}).
\emph{Editors / AEC chairs}: require a \textsc{swhid} for the \emph{Available} badge --- a reviewer
can verify it offline (Chapter~\ref{ch:repro}, with a ready call-for-artifacts box).
\emph{Funders / policy makers}: name Software Heritage and the \textsc{swhid} in conditions and
policy (Chapter~\ref{ch:outlook}, with a model clause and a governance/sustainability note).
\end{callout}

\newcommand{\pathcore}{\tikz[baseline=-0.5ex]\fill[swhaccent](0,0)circle(1.8pt);}
\newcommand{\pathskim}{\tikz[baseline=-0.5ex]{\draw[swhaccent,line width=0.4pt](0,0)circle(1.8pt);%
  \fill[swhaccent](0,0)--(90:1.8pt) arc(90:270:1.8pt)--cycle;}}
\newcommand{\pathskip}{\textcolor{swhdark!30}{\textperiodcentered}}

\clearpage
\begin{samepage}
\begin{callout}{How to read this note, by audience}
\small Not everyone needs every chapter. Pick a row and follow it; the rest can wait. The chapters
do build on one another --- the \textsc{swhid} (4) rests on the data model (3), and the practical
core (5) uses both --- so a skipped chapter is there when you need it.
\smallskip

\centering
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}l*{7}{c}c@{}}
\toprule
\textbf{Reader} & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} & \textbf{6} & \textbf{7} & \textbf{App.}\\
\midrule
Researcher \emph{(just do it)} & \pathcore & \pathskip & \pathskip & \pathskim & \pathcore & \pathskip & \pathskip & A\\
Reviewer / editor / AEC        & \pathcore & \pathskip & \pathskip & \pathcore & \pathskim & \pathcore & \pathskip & ---\\
Funder / policy maker          & \pathcore & \pathskip & \pathskip & \pathskip & \pathskip & \pathskip & \pathcore & E\\
Full course / student          & \pathcore & \pathcore & \pathcore & \pathcore & \pathcore & \pathcore & \pathcore & all\\
\bottomrule
\end{tabular}

\smallskip
{\footnotesize \pathcore~core \quad \pathskim~skim / optional \quad \pathskip~skip. \quad
\textbf{1}~Why \enspace \textbf{2}~Beyond FAIR \enspace \textbf{3}~Foundations \enspace
\textbf{4}~The \textsc{swhid} \enspace \textbf{5}~\textsc{ardc} \enspace
\textbf{6}~Reproducibility \enspace \textbf{7}~Policy.}
\end{callout}
\end{samepage}

\tableofcontents

% --- Chapters ------------------------------------------------------------
\include{chapters/01-why}
\include{chapters/02-beyond-fair}
\include{chapters/03-foundations}
\include{chapters/04-identifiers}
\include{chapters/05-ardc}
\include{chapters/06-reproducibility-aec}
\include{chapters/07-outlook}

% --- Appendices ----------------------------------------------------------
\appendix
\include{appendix/A-cookbook}
\include{appendix/B-swhid-syntax}
\include{appendix/C-glossary}
\include{appendix/D-swhid-tools}
\include{appendix/E-policy}

% --- References ----------------------------------------------------------
\printbibliography[heading=bibintoc,title={References}]

% --- Colophon: the note archives and verifies itself ---------------------
\input{colophon}

\end{document}
