\chapter{Software is not data --- and why ``beyond FAIR''}
\label{ch:beyond-fair}

\begin{objectives}\small
\begin{itemize}[nosep,leftmargin=*]
  \item explain how software differs from data: executable, evolving, dependency-laden;
  \item say where the FAIR principles fit software and where their \emph{I} and \emph{R}
        vocabulary misleads engineers;
  \item place the CODE roadmap and its single mandatory step.
\end{itemize}
\end{objectives}

The previous chapter asked us to treat research software as a first-class output. The natural
next question is: which recipe do we follow? For data, the community converged on a powerful
answer --- the FAIR principles~\autocite{fairdefinition2016} --- and it is tempting to reach for
the same recipe and simply declare software FAIR as well. This chapter argues, plainly, that the
temptation should be resisted: not because FAIR is wrong, but because software is not data, and
the parts of FAIR that matter most for software either come for free from archiving or mean
something quite different once code is involved.

\takeaway{FAIR is a starting point, not the destination, for living software.}

\section{The living nature of software}

Recall what source code is: knowledge that a machine can execute and a human can read. As Harold
Abelson put it, \enquote{programs must be written for people to read, and only incidentally for
machines to execute}; and as Len Shustek of the Computer History Museum observed, source code
offers \enquote{a view into the mind of the designer}. That double nature already sets software
apart from every other research output~\autocite{SwhCACM2018}.

Three further properties make software a moving target rather than a frozen artifact.

It \emph{evolves}. A dataset, once collected, is what it is; a piece of software that anyone
finds useful is a \emph{living entity}, continuously changing, with no state that can honestly be
called final~\autocite{NatureSD2026}. Its development history is not bookkeeping --- it is part
of the object's meaning, the record of how the design was reached.

It is \emph{diverse}. Behind the single word \enquote{software} hides a startling variety of
objects, differing in structure (a one-file script, or a million-line assembly of modules),
lifetime (a fire-and-forget analysis, or a library maintained for decades), community (one
researcher, one team, a worldwide crowd), size, and language --- written in any of the more than nine
thousand programming languages humanity has invented~\autocite{NatureSD2026}. A rule that fits a
tidy dataset will not fit all of this at once.

And it never stands \emph{alone}. Every non-trivial program rests on a biotope: a deep stack of
other software --- libraries, build tools, an operating system beneath it all. We have already
seen this (Figure~\ref{fig:matplotlib-deps}): one plotting library drags in a whole forest of
dependencies. In that biotope, the smallest change in a component buried far below can break the
reusability of the code on top --- or, worse, silently alter its
correctness~\autocite{NatureSD2026}. Data has no such problem; software has little else.

\section{FAIR for software: fit and friction}

The FAIR principles --- \emph{Findable, Accessible, Interoperable, Reusable} --- were proposed in
2016 for research \emph{data}~\autocite{fairdefinition2016}, and they have been enormously
successful there. In a nutshell, FAIR-for-data is about putting good metadata everywhere, which
is exactly the right obsession for data. Serious, well-intentioned efforts have since tried to
carry the same banner over to software: \emph{Towards FAIR principles for research
software}~\autocite{Lamprecht2020}, the FAIR4RS principles~\autocite{barker2022fair4rs}, and FAIR
computational workflows~\autocite{Goble2020}. We share their higher-level goal completely; the
disagreement is about \emph{how} --- and it is not a small one. Even when FAIR is carefully
adapted to software, it falls short of the problems one actually faces~\autocite{NatureSD2026}.

The friction has two sources. The first is that \enquote{FAIR} has drifted into a buzzword with
two faces: a vague, feel-good goal that policymakers invoke without looking inside, and a precise
set of recommendations that, for software, mostly reduce to \emph{putting metadata all over the
place}. Metadata is necessary, but nowhere near sufficient for code, and it is a trap to let FAIR
become the alpha and omega of how we treat software~\autocite{NatureSD2026}.

The second source of friction is sharper, and it is the one experienced developers feel at once.
The \emph{I} and the \emph{R} of FAIR --- \emph{interoperable} and \emph{reusable} --- are not
free-floating slogans in our field. They are precise, long-studied technical notions in software
engineering: interoperability is about APIs, ABIs, data formats, and protocols; reusability is
about modularity, portability, and architecture. These are properties that remain \emph{very hard
to achieve even for commercial software built by billion-dollar corporations}. When a researcher
who writes code is told to make her software \enquote{interoperable and reusable}, she hears a
decades-old engineering challenge --- not a request to fill in a metadata field. The FAIR
vocabulary, borrowed wholesale from data, thus confuses precisely the people best equipped to act.

\takeaway{For software, \emph{interoperable} and \emph{reusable} already mean something precise
--- and it is not what FAIR means by them.}

So what, in FAIR, do we keep? The first two letters. \emph{Findable} and \emph{accessible} are
real, achievable, and decisive for software --- and, happily, they are exactly what a universal
archive and an intrinsic identifier deliver, as the next two chapters show. The rest we should
not paper over with a checklist: software interoperability and reusability are a distinct, harder,
genuinely engineering problem, and pretending otherwise helps no one.

\begin{callout}{Objection, and response}
\small \emph{``Aren't you dismissing a serious community effort? FAIR4RS deliberately re-interprets
each principle \emph{for} software.''} --- Fairly put, and the FAIR4RS work~\autocite{barker2022fair4rs,Lamprecht2020}
is careful and well-meant; we share its goal of taking research software seriously. The disagreement
is narrow and specific: re-interpretation keeps a \emph{vocabulary} (interoperable, reusable) whose
established engineering meanings then mislead practitioners, and it leaves the load-bearing problems
(preservation, exact identity, provenance) to be solved by something \emph{else} --- which is exactly
what archiving and intrinsic identifiers are. We are not against FAIR's aims; we are for naming the
work by what it actually is. Read the FAIR4RS papers and judge for yourself.
\end{callout}

\section{The CODE-beyond-FAIR roadmap}

If imposing data principles on software is the wrong move, what is the right one? Stop legislating
from outside, and start learning from the community that has been building reusable software for
half a century: free and open-source software. Rather than minting new principles and handing them
to volunteers, we should adopt and adapt the philosophy that the FOSS world already lives
by~\autocite{NatureSD2026}.

That philosophy has to respect the diversity we just described. A single strict gold standard,
proclaimed from on high, would discourage exactly the people we most want to help --- the
scientist-developers who very often have no formal training in software
engineering~\autocite{Carver2022}, and who contribute, frequently, on goodwill alone. So the
recommendation is built, deliberately, as a \emph{progressive roadmap}: a short list of
\emph{mandatory} first steps, then a longer ladder of \emph{recommended} and \emph{optional}
ones, climbing toward a gold standard that nobody is required to reach all at once. The whole
roadmap fits a four-word mnemonic whose verbs --- \emph{Open, Document, Execute,
Collaborate} --- reshuffle into a word you will not forget: \textbf{CODE}~\autocite{NatureSD2026}.

\begin{description}[leftmargin=!,labelwidth=2.4cm,font=\bfseries\sffamily]
  \item[Open] publish the source on a public forge \emph{and} archive it in Software Heritage ---
        the two things the one \emph{mandatory} rung asks of you; then license it openly and claim authorship.
  \item[Document] meaningful names and comments first, then API documentation, examples, and
        notebooks.
  \item[Execute] declare your software and hardware dependencies, then ship a container or a Guix
        recipe, a test suite, and a worked example with its expected results.
  \item[Collaborate] handle issues, explain how to contribute, set honest limits, and grow a
        community.
\end{description}

\begin{figure}[htbp]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[font=\sffamily\small,
    rung/.style={draw=swhdark, rounded corners=1pt, minimum width=4.6cm, minimum height=0.9cm}]
    \node[rung, fill=swhaccent!85, text=white] (o) {\textbf{Open}\ \ {\footnotesize publish + archive}};
    \node[rung, fill=swhaccent!45, above right=2mm and 10mm of o] (d) {\textbf{Document}\ \ {\footnotesize docs, examples}};
    \node[rung, fill=swhaccent!22, above right=2mm and 10mm of d] (e) {\textbf{Execute}\ \ {\footnotesize deps, container, tests}};
    \node[rung, fill=swhbg, above right=2mm and 10mm of e] (c) {\textbf{Collaborate}\ \ {\footnotesize issues, community}};
    \node[below=1mm of o, font=\scriptsize\itshape] {mandatory};
    \node[above=1mm of c, font=\scriptsize\itshape] {optional};
  \end{tikzpicture}}
  \caption[The CODE roadmap]{The CODE roadmap as a ladder: only the first rung --- \textbf{Open}
  (publish the source and archive it in Software Heritage) --- is mandatory; \textbf{Document},
  \textbf{Execute}, and \textbf{Collaborate} are a progression you climb as far as your project
  warrants (the fill fades from mandatory to optional).}
  \label{fig:code}
\end{figure}

\takeaway{\textbf{CODE} --- Open, Document, Execute, Collaborate --- and only \emph{Open} is
mandatory.}

Above the mandatory rung, most of the work is simply choosing among well-worn tools --- none of them
required, all of them ordinary. Table~\ref{tab:toolbox} names the usual suspects for the
\textbf{Document} and \textbf{Execute} rungs; the point is not to use all of it, but to know the menu
exists.

\begin{table}[htbp]
  \centering\small
  \begin{tabularx}{\linewidth}{@{}>{\raggedright\arraybackslash}p{3.3cm} X@{}}
    \toprule
    \textbf{Concern} & \textbf{Common tools (pick what fits; none mandatory)} \\
    \midrule
    Version control & \texttt{git} --- used by more than nine in ten professional
      developers~\autocite{stackoverflow2022} --- with Conventional Commits, semantic versioning, and
      tagged releases that auto-archive to Zenodo and Software Heritage \\
    Build & Make, CMake, \texttt{cargo}, \texttt{poetry}, \texttt{npm} \\
    Scientific workflows & Snakemake, Nextflow, CWL \\
    Pinned environments & conda, pixi, uv; Nix, Guix; Docker, Apptainer \\
    Continuous integration & GitHub Actions, GitLab CI \\
    Executable papers & Jupyter notebooks served by BinderHub \\
    Reproducible builds & bit-for-bit determinism (\url{https://reproducible-builds.org/}) \\
    \bottomrule
  \end{tabularx}
  \caption[A toolbox for the CODE rungs]{A sampling of the tooling behind the \textbf{Document} and
  \textbf{Execute} rungs of CODE. None of it is mandatory --- the only required step remains
  \textbf{Open} (publish and archive) --- but knowing which tools exist is half the battle.}
  \label{tab:toolbox}
\end{table}

Notice where the single hard requirement lands: \emph{publish and archive}. Everything above it is
a ladder you climb as far as your project warrants. This is liberating, not lax --- it says, in
effect, that the worst outcome is not imperfect software but \emph{lost} software, and that
\enquote{ugly code is always better than no code}~\autocite{Leveque2013}. It also tells us where
these notes go next. The mandatory rung of \textbf{O}pen --- archive your code, and be able to
point at it --- is precisely the \textbf{A}rchive and \textbf{R}eference of the \textsc{ardc} practice at
the heart of these notes. To make good on it we need two things first: somewhere to keep code for
the long term, and a way to name exactly the code we mean. Those are the next two chapters.
