\chapter{Glossary}
\label{app:glossary}

\begin{description}[leftmargin=!,labelwidth=3.1cm,font=\bfseries,style=nextline]
  \item[Anchor] a context qualifier of a \textsc{swhid}: the node (\texttt{dir}/\texttt{rev}/%
        \texttt{rel}/\texttt{snp}) against which a \texttt{path} is resolved.
  \item[Archive] a system whose purpose is to \emph{keep} content for the long term --- as opposed
        to a forge, which is built to \emph{develop} it.
  \item[ARDC] Archive, Reference, Describe, Cite \& credit --- the four operations that make up the
        practice at the heart of these notes.
  \item[Artifact evaluation] the process, now standard at computer-science conferences, of checking
        that the software behind a paper exists, runs, and supports its claims.
  \item[Badge (ACM)] one of the ACM \emph{Artifact Review and Badging} marks: Artifacts Available;
        Artifacts Evaluated --- Functional / Reusable; Results Reproduced / Replicated.
  \item[\texttt{biblatex-software}] the \LaTeX{} bibliography style for citing software, adding four
        entry types and a \texttt{swhid=} field.
  \item[CITATION.cff] the Citation File Format, a lightweight YAML file describing how to cite a
        repository.
  \item[CODE] the beyond-FAIR roadmap for research software: \textbf{O}pen, \textbf{D}ocument,
        \textbf{E}xecute, \textbf{C}ollaborate.
  \item[CodeMeta] a JSON-LD vocabulary for software metadata, written to a \texttt{codemeta.json}.
  \item[Content (\texttt{cnt})] the bytes of a single file; the leaf of the Merkle DAG.
  \item[Core SWHID] \swhid{swh:1:\textit{type}:\textit{hash}}; the part that \emph{identifies}, a
        fingerprint of the content alone.
  \item[Deposit] the \textsc{sword}-based archival channel, for trusted partners and registered
        accounts, used to push code that is not in a public forge.
  \item[Directory (\texttt{dir})] a directory-tree node: names, permissions, and the hashes of its
        children.
  \item[DOI] Digital Object Identifier; an \emph{extrinsic} identifier resolved through the Handle
        System.
  \item[Extrinsic identifier] a name assigned by a registry; it means nothing without that registry.
  \item[FAIR] the Findable, Accessible, Interoperable, Reusable principles --- designed for research
        \emph{data}.
  \item[FAIR4RS] the effort to adapt the FAIR principles to research software.
  \item[Forge] a collaborative development platform (GitHub, GitLab, Bitbucket, \dots) --- \emph{not}
        an archive.
  \item[Intrinsic identifier] a name computed from the content itself; verifiable offline, with no
        registry and no middle-man.
  \item[Lister / loader] the components that, respectively, enumerate the repositories on a hosting
        platform and ingest each one into the archive.
  \item[Merkle DAG] the content-addressed directed acyclic graph in which Software Heritage stores
        all code and history; each node is named by the hash of everything beneath it.
  \item[Origin] a context qualifier: the URI a piece of code was harvested from.
  \item[Path] a context qualifier: an absolute path from the anchor's root directory.
  \item[Qualified SWHID] a core \textsc{swhid} plus context (\texttt{origin}, \texttt{visit},
        \texttt{anchor}, \texttt{path}) and fragment (\texttt{lines}, \texttt{bytes}) qualifiers.
  \item[Release (\texttt{rel})] a named release node (an annotated tag).
  \item[REUSE] a specification for declaring per-file licensing in a machine-readable way.
  \item[Revision (\texttt{rev})] a commit node --- identical to the underlying Git commit hash.
  \item[Save Code Now] the web entry point (\url{https://save.softwareheritage.org}) that triggers
        archival of a public repository.
  \item[Snapshot (\texttt{snp})] the full state of a repository at one visit --- all of its branches
        and tags.
  \item[SPDX] the standard catalogue of machine-readable software-licence identifiers.
  \item[SWHID] \emph{Software Hash Identifier} --- the intrinsic identifier for source code
        (ISO/IEC~18670:2025). Originally read \enquote{Software Heritage Identifier}; since the
        specification it is \enquote{Software \emph{Hash} Identifier}.
  \item[Visit] a crawl of an origin by the archive; also the context qualifier giving that visit's
        snapshot \textsc{swhid}.
\end{description}
